---
title: "Coming from ggplot2"
description: >
  A translation guide for ggplot2 users: the same plots built both ways,
  side by side, and the handful of concepts that differ.
---

```{r}
#| include: false
have_all <- all(vapply(
  c("vellum", "vellumplot", "vellumwidget"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
if (have_all) suppressMessages(library(vellumverse))
```

If you already think in `ggplot2`, `vellumplot` will feel familiar: data in,
aesthetics mapped to marks, scales and facets and themes layered on top. The
grammar is the same idea. Four things are different, and once they click the
rest is a matter of renaming.

Throughout, the `ggplot2` code is shown for reference and the `vellumplot`
equivalent is the one rendered.

## The four differences

**1. `|>`, not `+`.** Layers are joined with the base pipe. A `vellumplot` plot is
an ordinary pipeline, so the first argument of every verb is the plot.

**2. No `aes()`.** Aesthetics are named arguments on the mark itself. A bare
column name is mapped to data; a literal value is a constant. `color = hp`
is a channel with a legend; `color = "red"` paints everything red.

**3. `geom_*` becomes `mark_*`.** Same layers, different prefix.
`geom_point()` is `mark_point()`, `geom_smooth()` is `mark_smooth()`, and so
on.

**4. A plot is a spec, not a picture.** `vellumplot` builds an inspectable
specification and compiles it only when printed, saved, or made interactive.
`summary()` shows you the spec without drawing it.

## Translation table

| ggplot2 | vellumplot |
|---|---|
| `ggplot(df, aes(x, y))` | `vplot(df)` then `mark_*(x = x, y = y)` |
| `aes(color = z)` | `color = z` on the mark |
| `+` | `\|>` |
| `geom_point()` | `mark_point()` |
| `geom_line()` / `geom_path()` | `mark_line()` |
| `geom_col()` / `geom_bar()` | `mark_bar()` |
| `geom_smooth()` | `mark_smooth()` |
| `geom_histogram()` | `mark_histogram()` |
| `geom_boxplot()` | `mark_boxplot()` |
| `geom_tile()` / `geom_raster()` | `mark_tile()` / `mark_raster()` |
| `geom_hex()` | `mark_hex()` |
| `geom_sf()` | `mark_sf()` |
| `scale_color_continuous()` | `scale_color_continuous()` |
| `scale_*_manual(values = )` | `scale_*_manual(values = )` |
| `facet_wrap(~z)` | `facet_wrap(~z)` |
| `facet_grid(a ~ b)` | `facet_grid(a ~ b)` |
| `coord_flip()` / `coord_polar()` | `coord_flip()` / `coord_polar()` |
| `theme_minimal()` | `theme_minimal()` |
| `labs(title = )` | `labs(title = )` |
| `ggsave("p.png")` | `render_plot(p, "p.png")` |

Note that spelling: `vellumplot` uses American `color` as the primary aesthetic
name (with `colour`/`fill` accepted where it makes sense).

## Scatter with a colour scale

::: {layout-ncol=2}
```r
# ggplot2
ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point() +
  scale_color_continuous()
```

```r
# vellumplot
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```
:::

```{r scatter}
#| echo: false
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```

## Points with a fitted line

The layering translates one-to-one; the `+` becomes a `|>` and each geom
becomes a mark. Note that `vellumplot` marks carry their own encodings rather than
inheriting a top-level `aes()`.

::: {layout-ncol=2}
```r
# ggplot2
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```r
# vellumplot
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```
:::

```{r smooth}
#| echo: false
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

## Faceting

`facet_wrap()` and `facet_grid()` keep their names and their formula
interface.

::: {layout-ncol=2}
```r
# ggplot2
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point() +
  facet_wrap(~cyl) +
  theme_minimal()
```

```r
# vellumplot
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  facet_wrap(~cyl) |>
  theme_minimal()
```
:::

```{r facet}
#| echo: false
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  facet_wrap(~cyl) |>
  theme_minimal()
```

## A bar chart with a manual palette

::: {layout-ncol=2}
```r
# ggplot2
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  scale_fill_manual(values = c("#6b4f2c", "#8a7350")) +
  coord_flip()
```

```r
# vellumplot
vplot(mpg) |>
  mark_bar(x = class, fill = class) |>
  scale_fill_manual(values = c("#6b4f2c", "#8a7350")) |>
  coord_flip()
```
:::

## What you gain

Two things `ggplot2` cannot easily do fall out of the spec-first design.

**Inspect a plot before drawing it.** Because the plot is data, `summary()`
prints its structure:

```{r}
#| eval: false
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  summary()
#> <PlotSpec> 32x11 (11 columns), page 6x4 in
#>
#> ── layers
#> • mark_point(x = wt, y = mpg, color = hp)
```

**Make it interactive for free.** The same spec, handed to `as_widget()`,
becomes an HTML widget. The interactive channels (`tooltip`, `data_id`) are
more mark arguments:

```{r}
#| eval: false
df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg, model = rownames(mtcars))

vplot(df) |>
  mark_point(x = wt, y = mpg, tooltip = model, data_id = model) |>
  as_widget()
```

## What is different by design

A few `ggplot2` habits do not carry over, on purpose:

- There is no global `aes()` shared across layers; each mark states its own
  encodings. This is more verbose for multi-layer plots but removes the
  inheritance rules you have to keep in your head.
- `+` chaining and its operator methods are gone; everything is a function
  in a pipe.
- Saving is `render_plot()`, not `ggsave()`, and one spec renders to PNG,
  SVG, and PDF from the same object. See [one scene, three
  outputs](one-scene-three-outputs.html).

For the exhaustive list of marks, scales, coords, and themes, see the
[vellumplot reference](https://r-vellum.github.io/vellumplot/).
