---
title: "The vellum ecosystem"
description: >
  How vellum, quill, and gloss fit together — from a low-level scene graph,
  to a grammar of graphics, to an interactive widget.
---

```{r, include = FALSE}
have_all <- all(vapply(
  c("vellum", "quill", "gloss"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/ecosystem-",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
```

`scriptorium` is a meta-package: it holds no plotting code of its own. Its
job is to install and attach the three packages that make up the *vellum*
graphics ecosystem, so that

```{r setup}
library(scriptorium)
```

puts all of them on your search path at once. Each one owns a distinct layer
of the stack, and the design intent is that you rarely think about the
boundaries between them.

## Three layers, one scene

The metaphor is a medieval **scriptorium** — the room where manuscripts were
written. The parchment, the pen, and the annotation each have a name:

| layer | package | what it is | the tidyverse analogue |
|---|---|---|---|
| backend | **vellum** | a retained scene graph, unit/layout engine, and PNG/SVG/PDF renderer, with a Rust engine | `grid` |
| grammar | **quill** | a pipe-first grammar of graphics that *compiles* a plot spec into a vellum scene | `ggplot2` |
| interaction | **gloss** | turns a vellum scene into a self-contained interactive HTML widget | `htmlwidgets` |

The important word is **compiles**. A `quill` plot is not drawn as you build
it; it is an inspectable specification that is turned into a `vellum` scene
only when it is printed, rendered, or handed to `gloss`. Everything below is
therefore one scene passing down the stack.

## vellum — the parchment

`vellum` is the foundation. You describe a scene with a small declarative
API and render it; the scene graph, layout, and drawing run in Rust, and the
same scene renders to raster or vector.

```{r vellum}
vl_scene(width = 4, height = 2.4, bg = "white") |>
  draw(rect_grob(
    width = 0.9, height = 0.8,
    gp = gpar(fill = linear_gradient(c("#6b4f2c", "#c9a86a")), col = NA)
  )) |>
  draw(text_grob(
    "vellum", x = 0.5, y = 0.5,
    gp = gpar(fontsize = 44, col = "white", fontface = "bold")
  ))
```

You would reach for `vellum` directly to build a bespoke visual, or as the
target of a higher-level tool — which is exactly what `quill` is.

## quill — the pen

`quill` gives you the grammar: data in, marks and scales declared with tidy
evaluation, everything trained and laid out for you.

```{r quill}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```

Because the plot is a spec, you can inspect it without drawing anything:

```{r quill-summary, eval = FALSE}
summary(vplot(mtcars) |> mark_point(x = wt, y = mpg, color = hp))
#> <PlotSpec> 32x11 (11 columns), page 6x4 in
#>
#> ── layers
#> • mark_point(x = wt, y = mpg, color = hp)
```

## gloss — the annotation

`gloss` is terminal: hand it a `quill` plot (or a bare `vellum` scene) and it
compiles the scene, emits the SVG plus a per-element table, and returns an
htmlwidget with hover tooltips, selection, brushing, and pan/zoom — no Shiny,
no server.

```{r gloss, eval = FALSE}
df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg, model = rownames(mtcars))

vplot(df) |>
  mark_point(x = wt, y = mpg, tooltip = model, data_id = model) |>
  as_widget()
```

The `tooltip` and `data_id` arguments are declared in `quill`, read by
`gloss`, and carried through the `vellum` scene — the three layers agreeing
on one contract.

## Where each package's docs live

`scriptorium` is the front door; each package documents itself in full:

- [vellum](https://schochastics.github.io/vellum/) — grobs, units, viewports,
  the paint model, `datashade()`, and grid interop.
- [quill](https://schochastics.github.io/quill/) — the full mark, scale,
  facet, coord, and theme reference.
- [gloss](https://schochastics.github.io/gloss/) — interaction options,
  linked views, and crosstalk interop.
