---
title: "Extending the ecosystem"
description: >
  The seams the three layers agree on, and how to build on them: composite
  grobs, custom themes, the interactivity contract, and teaching the ecosystem
  about your own objects.
---

```{r}
#| include: false
have_all <- all(vapply(
  c("vellum", "quill", "gloss"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
if (have_all) suppressMessages(library(scriptorium))
```

The ecosystem is layered so that each layer exposes a seam the next one
builds on. The same seams are open to you. This article walks the four most
useful extension points, from the most local (a reusable piece of drawing) to
the most structural (making the whole ecosystem understand a new kind of
object).

## 1. Reusable drawing: grob factories and scene helpers

There are two ways to package a piece of drawing for reuse, and both are
ordinary R functions — nothing to register, nothing to teach the renderer.

The first is a **grob factory**: a function that returns a single grob.
`datashade()` is exactly this — it returns one `raster_grob` — and a function
that returns a grob is a first-class primitive that drops straight into
`draw()`, inherits graphical parameters from its viewport, and renders to
every backend.

The second, for drawing that is more than one primitive, is a **scene
helper**: a function that takes a scene, draws into it, and returns the scene.
Because `draw()` takes one grob at a time, this is how you compose several
primitives into a reusable unit that still chains in a pipe:

```{r composite}
# a labelled marker: a filled circle with text centred on it
marker <- function(scene, x, y, label, fill = "#6b4f2c") {
  scene |>
    draw(circle_grob(x = x, y = y, r = 0.06, gp = gpar(fill = fill, col = NA))) |>
    draw(text_grob(label, x = x, y = y,
                   gp = gpar(col = "white", fontface = "bold", fontsize = 12)))
}

vl_scene(4, 2, bg = "white") |>
  marker(0.25, 0.5, "A") |>
  marker(0.5, 0.5, "B", fill = "#8a7350") |>
  marker(0.75, 0.5, "C")
```

Scene-in, scene-out means your helper composes with `draw()`, `push()`, and
`pop()` like any built-in verb.

## 2. Custom themes: restyle without new machinery

At the grammar layer, a theme is a function that sets styling on a plot spec.
`quill` gives you `theme()` for individual elements and `set_theme()` for the
common surfaces, built from `element_text()`, `element_line()`, and
`element_rect()`. A reusable theme is a function that applies them:

```{r theme}
theme_parchment <- function(plot) {
  plot |>
    set_theme(
      panel_bg  = "#faf5ea",
      grid_col  = "#e6dcc4",
      label_col = "#4a371f"
    ) |>
    theme(
      plot.title     = element_text(family = "serif", size = 16),
      legend.position = "bottom"
    )
}

vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  theme_parchment()
```

Since a theme is an ordinary verb in the pipe, it composes with everything
else and can be shared as a one-function package.

## 3. The interactivity contract

This is the seam that ties all three layers together, and the most useful one
to understand if you build tools on top. Interactivity is not a feature of
`gloss` alone; it is a *contract* carried through the scene:

- A `quill` mark declares the reserved channels `tooltip`, `data_id`,
  `hover_group`, `hover_color`, and `selected_color`.
- Those are compiled onto the `vellum` grobs as their `key` (identity) and
  `meta` (the rest).
- `gloss` reads them back out of the compiled scene — via `scene_model()` —
  and wires up the widget.

The practical consequence: **you can make a bespoke `vellum` scene
interactive without touching `gloss` at all**, just by attaching `key` and
`meta` to your grobs. Every primitive grob accepts them.

```{r}
#| eval: false
# a hand-built scene whose points are hoverable and selectable
s <- vl_scene(4, 3, bg = "white") |>
  push(viewport(xscale = c(0, 10), yscale = c(0, 10))) |>
  draw(points_grob(
    x = unit(c(2, 5, 8), "native"),
    y = unit(c(3, 7, 4), "native"),
    gp = gpar(fill = "#6b4f2c"),
    key = c("a", "b", "c"),                     # identity for selection
    meta = list(tooltip = c("Alpha", "Beta", "Gamma"))
  ))

as_widget(s)   # tooltips and selection work, no quill involved
```

Because the contract lives in the scene, a mark you invent in `quill` that
sets `tooltip`/`data_id` becomes interactive in `gloss` automatically, and a
grob you build in `vellum` that sets `key`/`meta` does too. Neither end has
to know about the other.

## 4. Teach the ecosystem about your object

The deepest seam is `as_vellum_scene()`, the S7 generic that every consumer
calls to turn *something* into a renderable scene. `quill` registers a method
for its plot spec; that is how `render()` and `as_widget()` accept a `quill`
plot at all. If you have your own object — a model, a spatial structure, a
domain-specific plot type — you can register a method that compiles it into a
`vellum` scene, and from that moment it renders and becomes interactive
through the same machinery as everything else.

```{r}
#| eval: false
library(S7)

# a tiny custom object
my_plot <- function(data) structure(list(data = data), class = "my_plot")

# teach the ecosystem how to compile it into a vellum scene
method(as_vellum_scene, new_S3_class("my_plot")) <- function(x, ...) {
  vl_scene(4, 3, bg = "white") |>
    push(viewport(xscale = range(x$data$x), yscale = range(x$data$y))) |>
    draw(points_grob(unit(x$data$x, "native"), unit(x$data$y, "native"),
                     gp = gpar(fill = "#6b4f2c")))
}

p <- my_plot(data.frame(x = 1:10, y = (1:10)^2))

render(p, "custom.png")   # works: render() compiles via as_vellum_scene()
as_widget(p)              # works: gloss compiles via the same seam
```

Registering that one method is what makes `render()`, `render_plot()`, and
`as_widget()` all accept your object — they never see `my_plot`, only the
scene it compiles to.

## The picture

The four seams line up with the four layers of reuse:

| you want to reuse | the seam | where it lives |
|---|---|---|
| a piece of drawing | a function returning grobs | `vellum` |
| a look | a function applying `theme()`/`set_theme()` | `quill` |
| interactivity | `key`/`meta` on grobs; `tooltip`/`data_id` on marks | the scene contract |
| a whole new plot type | an `as_vellum_scene()` method | the compiler seam |

Each is independent: a composite grob needs no theme, a theme needs no custom
object, and a custom object gets rendering and interactivity for free the
moment it can produce a scene. That independence is the point of splitting the
ecosystem into layers — see [the vellum ecosystem](ecosystem.html) and
[vellum's design principles](design-principles.html) for the reasoning behind
the split.
