---
title: "Getting started"
description: >
  From a data frame to a rendered plot to an interactive widget, in one
  pipeline, using the whole vellum ecosystem through vellumverse.
---

```{r}
#| include: false
have_all <- all(vapply(
  c("vellum", "vellumplot", "vellumwidget"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
```

This is the short path from a data frame to a finished plot. The
[ecosystem article](ecosystem.html) explains *why* the pieces are split the
way they are; here we just use them.

## Install and attach

The ecosystem compiles a Rust crate (inside `vellum`), so you need a Rust
toolchain (`cargo`/`rustc`) on your machine alongside R. With that in place:

```{r}
#| eval: false
# install.packages("pak")
pak::pak("r-vellum/vellumverse")
```

`vellumverse` is a meta-package. Attaching it attaches the three packages
that do the real work, so a single call puts the whole grammar on your
search path:

```{r setup}
library(vellumverse)
```

You now have `vplot()` and the marks (from `vellumplot`), the scene and rendering
primitives (from `vellum`), and `as_widget()` (from `vellumwidget`).

## Your first plot

Every plot starts with `vplot()` on a data frame, then adds a *mark*. A mark
maps columns to visual channels — position, colour, size — and the arguments
are ordinary column names, evaluated in the data:

```{r first}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg)
```

There is no `aes()` and no `+`. Aesthetics are named arguments on the mark,
and layers are joined with the base pipe `|>`.

## Map a column to colour

Give an aesthetic a column and it becomes a *channel* with a legend; give it
a bare value and it becomes a constant. Here `color = hp` is a channel,
trained into a continuous scale:

```{r color}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```

Compare with a fixed colour, which draws every point the same and adds no
legend:

```{r constant}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = "#6b4f2c")
```

The rule is worth remembering: a symbol or expression (`hp`, `factor(cyl)`)
is data; a literal (`"#6b4f2c"`, `3`) is a constant.

## Add layers

Marks stack. Add a fitted line over the points by piping in another mark:

```{r layers}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

## Facet, then theme

`facet_wrap()` splits the plot into small multiples by a variable, and a
`theme_*()` restyles the whole thing:

```{r facet}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  facet_wrap(~cyl) |>
  theme_minimal()
```

## Inspect without drawing

A `vellumplot` plot is a *spec*: an object describing the plot, not a picture.
Nothing is drawn until you print or render it, so you can examine what you
have built with `summary()`:

```{r}
#| eval: false
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  summary()
#> <PlotSpec> 32x11 (11 columns), page 6x4 in
#>
#> ── layers
#> • mark_point(x = wt, y = mpg, color = hp)
```

## Save to a file

`render_plot()` writes the plot to disk. The format follows the file
extension — raster for `.png`, vector for `.svg` and `.pdf`:

```{r}
#| eval: false
p <- vplot(mtcars) |> mark_point(x = wt, y = mpg)

render_plot(p, "cars.png")   # raster
render_plot(p, "cars.pdf")   # vector
```

The same spec produces every format; see [one scene, three
outputs](one-scene-three-outputs.html) for why that is more than a
convenience.

## Make it interactive

Hand the same pipeline to `as_widget()` and it becomes a self-contained
HTML widget — hover tooltips, click selection, brushing, and pan/zoom, with
no Shiny and no server. The interactive bits are declared on the mark:
`tooltip` is what a point shows on hover, `data_id` is its identity for
selection.

```{r}
#| eval: false
df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg, model = rownames(mtcars))

vplot(df) |>
  mark_point(x = wt, y = mpg, color = mpg, tooltip = model, data_id = model) |>
  scale_color_continuous() |>
  as_widget()
```

That is the whole arc: the same spec you printed, saved as a PDF, and turned
into a widget, without rewriting it for each destination.

## Where to go next

- [Coming from ggplot2](coming-from-ggplot2.html) — a translation table if
  you already think in `geom_*` and `aes()`.
- [The vellum ecosystem](ecosystem.html) — how the three layers divide the
  work.
- Each package's own site has the full reference:
  [vellumplot](https://r-vellum.github.io/vellumplot/) for every mark, scale, and
  theme; [vellum](https://r-vellum.github.io/vellum/) for the drawing
  primitives; [vellumwidget](https://r-vellum.github.io/vellumwidget/) for interaction
  options.
