---
title: "One scene, three outputs"
description: >
  The ecosystem's central move: a plot is compiled once into a vellum scene,
  then rendered to raster, to vector, or to an interactive widget from the
  same object.
---

```{r}
#| include: false
have_all <- all(vapply(
  c("vellum", "quill", "gloss"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
if (have_all) suppressMessages(library(scriptorium))
```

Most plotting workflows re-specify the plot for each destination: one code
path for the PNG in a report, another for the vector figure a journal wants,
a third library entirely if the plot needs to be interactive on the web. The
vellum ecosystem is built so that the *specification* and the *destination*
are separate concerns. You describe the plot once; where it ends up is a
choice made at the end.

## The seam

A `quill` plot is a spec. It is compiled into a `vellum` scene — a concrete,
laid-out tree of drawing primitives — through a single generic,
`as_vellum_scene()`. Everything downstream consumes that scene:

- `vellum`'s `render()` writes the scene to a file, choosing raster or vector
  from the extension.
- `gloss`'s `as_widget()` turns the same scene into an HTML widget.

Both call `as_vellum_scene()` under the hood. That one seam is why the same
object can go three places without being rewritten.

## Build the spec once

```{r spec}
df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg, model = rownames(mtcars))

p <- vplot(df) |>
  mark_point(x = wt, y = mpg, color = mpg, tooltip = model, data_id = model) |>
  scale_color_continuous() |>
  labs(title = "Fuel economy")
```

Nothing has been drawn. `p` is a description — you can inspect it, pass it
around, store it, and decide its fate later:

```{r}
#| eval: false
summary(p)
#> <PlotSpec> 32x3 (3 columns), page 6x4 in
#>
#> ── layers
#> • mark_point(x = wt, y = mpg, color = mpg)
```

Notice that `tooltip` and `data_id` are already on the mark. They mean
nothing to a static render and everything to the widget — carrying them in
the spec from the start is what lets one object serve both.

## Output 1 — raster

Printing the spec draws it into the active device; that is what the chunk
below does, and it is the same path a raster file takes.

```{r raster}
p
```

To write it to disk as a PNG:

```{r}
#| eval: false
render_plot(p, "fuel.png")          # tiny-skia raster
render_plot(p, "fuel.png", dpi = 300)  # override the authored resolution
```

## Output 2 — vector

The identical spec, sent to a vector extension, is rendered through a
different backend — a hand-rolled SVG writer, or krilla for PDF — with no
change to the plot code:

```{r}
#| eval: false
render_plot(p, "fuel.svg")   # vector SVG
render_plot(p, "fuel.pdf")   # vector PDF
```

Because the layout is computed once and then walked for each backend, the
raster and vector versions are the same plot, not two approximations of it —
the point positions, text metrics, and legend placement are identical.

## Output 3 — interactive widget

The same `p`, handed to `as_widget()`, compiles to the same scene and then
emits an SVG plus a per-element table, wired up with hover, selection,
brushing, and pan/zoom — client-side, no Shiny:

```{r}
#| eval: false
as_widget(p)
```

The `tooltip` and `data_id` you declared back in the spec are what the widget
reads. Three layers agreeing on one contract: `quill` declares the channels,
`vellum` carries them through the scene, `gloss` acts on them.

## Why this matters

The payoff is not just fewer keystrokes. Because there is one spec and one
compiled scene:

- **The outputs cannot drift.** A raster figure and its interactive
  counterpart are guaranteed to show the same plot, because they *are* the
  same scene.
- **The decision is deferred.** You can write analysis code that returns a
  spec and let the calling context — a Quarto chunk, a Shiny app, a batch
  job writing PDFs — decide how to realise it.
- **Bare `vellum` scenes get the same treatment.** `as_widget()` and
  `render()` accept a hand-built `vellum` scene too, so a bespoke visual made
  with grobs travels the same three roads as a grammar plot.

## Going lower: a bare vellum scene

The same three-way rendering exists one layer down, without the grammar. A
`vellum` scene renders to all three formats directly:

```{r}
#| eval: false
s <- vl_scene(4, 3) |>
  draw(circle_grob(r = 0.3, gp = gpar(fill = "tomato", col = NA)))

render(s, "out.png")   # raster   (tiny-skia)
render(s, "out.svg")   # vector   (hand-rolled SVG)
render(s, "out.pdf")   # vector   (krilla)
as_widget(s)           # interactive widget
```

The grammar in `quill` is, in the end, a convenient way to produce one of
these scenes. Whether the scene came from `vplot()` or from raw grobs, the
destinations are the same.

## See also

- [The vellum ecosystem](ecosystem.html) — the layers and how they connect.
- [Performance and big data](performance.html) — what compiling once buys you
  when the data is large.
