---
title: "Performance and big data"
description: >
  Why the ecosystem stays fast as data grows, and how datashade renders
  millions of points by aggregating before it draws.
---

```{r}
#| include: false
have_all <- all(vapply(
  c("vellum", "vellumplot", "vellumwidget"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = have_all
)
if (have_all) suppressMessages(library(vellumverse))
```

Two things make plotting slow at scale: computing the layout, and drawing
one graphical object per data point. The vellum ecosystem attacks both. The
scene graph, layout, and rendering run in a Rust engine rather than in R, and
for the case where the data has more points than the screen has
pixels, `datashade()` aggregates the data into an image *before* anything is
drawn.

## Where the time goes, and doesn't

In a `grid`-based stack, a plot's display list is replayed on every resize,
text has to be measured against whatever device is current, and each point is
a grob the engine walks at draw time. `vellum` moves that work down into
Rust: metrics are available at construction time, layout is a single explicit
pass, and one render walk emits the chosen backend. The consequence for
performance is that the R-side cost of a plot is building the spec, which is
cheap; the drawing cost is paid once, in compiled code.

That handles moderately large plots well. But no amount of engine speed helps
if the plot asks the renderer to draw two million individual circles into an
SVG: the file alone would be enormous, and most of those marks would land on
the same pixel. That is the problem `datashade()` solves.

## datashade: aggregate, then draw

`datashade()` bins points into a grid the size of the output image, counts
how many land in each cell, and colours the cells by density. It returns a
single raster, one image, instead of millions of marks. The binning is a
one-pass aggregation in Rust, so the cost scales with the number of *pixels*,
not the number of *points*. Ten thousand points and ten million points cost
almost the same to draw, because the output is the same-sized image either
way.

### The grammar-level way

In `vellumplot`, `mark_datashade()` is a mark like any other. Here are a million
points, drawn as a density field:

```{r datashade-mark}
set.seed(1)
n <- 1e6
df <- data.frame(
  x = rnorm(n),
  y = rnorm(n) * 0.5 + rnorm(n)
)

vplot(df) |>
  mark_datashade(x = x, y = y)
```

`mark_datashade()` takes `width` and `height` (the aggregation grid, default
`400 × 300`), a `colors` ramp, and `how`, which controls how counts map to
colour. The default `how = "eq_hist"` equalises the histogram so that a few
very dense cells do not wash out the rest of the structure.

### The backend-level way

The same aggregation is available in `vellum` as `datashade()`,
which returns a `raster_grob` you place in a scene yourself. Use this when
you are building a bespoke visual rather than a grammar plot:

```{r datashade-vellum}
set.seed(1)
n <- 1e6
x <- rnorm(n)
y <- x * 0.5 + rnorm(n)

vl_scene(6, 4.5, bg = "white") |>
  push(viewport(xscale = range(x), yscale = range(y))) |>
  draw(datashade(x, y, width = 450, height = 300,
                 colors = c("#fde0dd", "#7a0177")))
```

Because `datashade()` bins over `xlim × ylim` (defaulting to the data range),
you draw it inside a viewport whose `xscale`/`yscale` match those limits so
the image lines up with axes you add around it.

## Tuning the density mapping

`how` decides how cell counts become colour, and it is the knob that most
changes what you see:

- `"eq_hist"` (default): histogram equalisation; best for revealing
  structure across a wide range of densities.
- `"log"`: log of the count; a gentler compression than equalisation.
- `"cbrt"`: cube root; gentler still.
- `"linear"`: raw counts; only useful when density is fairly uniform.

`colors` is any vector of colours forming the ramp from sparse to dense, and
`weight` lets you accumulate a summed quantity per cell instead of a plain
count, a weighted density.

```{r}
#| eval: false
# emphasise the tails with log scaling, custom ramp
vplot(df) |>
  mark_datashade(x = x, y = y, how = "log",
                 colors = c("#f7fbff", "#08306b"))
```

## When to reach for it

The rule of thumb is the pixel count. If your scatter has more points than
the plot has pixels, tens of thousands and up, overplotting means most
marks are invisible anyway, and `datashade()` shows the density they were
hiding while rendering in constant time. Below that, a plain `mark_point()`
is fine and keeps every point individually addressable (which matters for
interactivity; a datashaded image is one raster, not a thousand hoverable
points).

## Choosing an output format at scale

The format you render to interacts with data size. A raster target (`.png`)
is a fixed-size grid of pixels regardless of how much data went into it. A
vector target (`.svg`, `.pdf`) stores every mark as an element, so a
half-million-point scatter produces a half-million-element file. For large
data, prefer raster output, or datashade first so that even the vector output
is a single embedded image. See [one scene, three
outputs](one-scene-three-outputs.html) for how the same spec reaches each
format.

## Diagnosing a slow or heavy scene

`vellum` ships a couple of tools for when a plot is larger or slower than you
expect:

```{r}
#| eval: false
why_size(s)             # explain what is driving a scene's rendered size
vl_clear_render_cache() # drop cached render state
```

For scenes with an expensive sub-tree that is drawn repeatedly, a viewport
can cache its rendered content with `viewport(..., cache = TRUE)`, so the work
is done once and reused.

## Summary

- The engine (scene graph, layout, render) runs in Rust, so per-plot R
  overhead is building the spec.
- `datashade()` / `mark_datashade()` render arbitrarily many points in time
  proportional to the image size, by aggregating to a density raster before
  drawing.
- `how` and `colors` shape how density reads; `weight` gives weighted
  densities.
- For large data, prefer raster output or datashade before going to vector.
